---
title: Explore the effects of major storms and weather events on public health and
  economic consequences in the United States
author: "Dominic Tran"
output: 
  html_document:
    keep_md: true
---


## Synopsis
In this report, the aim is to explore and answer simple questions: (1) which weather event types are most harmful to the United States population, (2) which weather event types have the most economic consequences across the United States.

The data used in this study is based on the U.S. National Oceanic and Atmospheric Administration's (NOAA) [storm database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

As can be seen in the report, Tornado/Waterspout is the most harmful to 
the United States population; and Flood/Small Stream causes the most financial damage.

## Data Processing
```{r library}
library(ggplot2)
```

Download and import data
```{r cache=TRUE}
# unfortunate - allocated error was reported for the below code
#data.zip <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
#temp <- tempfile()
#download.file(data.zip, temp, method="curl")
#data <- read.csv(bzfile(temp), na.strings = "?", stringsAsFactors = FALSE)
#unlink(temp)
```

To work around, assuming that "repdata-data-StormData.csv.bz2" is in the current
working directory
```{r cache=TRUE}
data.zip <- "repdata-data-StormData.csv.bz2"
data <- read.csv(bzfile(data.zip), na.strings = "?", stringsAsFactors = FALSE)
```

Clean data
```{r}
data <- data[!is.na(data$EVTYPE),]
data$EVTYPE <- toupper(data$EVTYPE)
data <- data[grep("SUMMARY", data$EVTYPE, invert = TRUE),]
data <- data[grep("NONE", data$EVTYPE, invert = TRUE),]
```

Prepare for transformation
Mapping of "EVTYPE" to related/similar weather event types.
```{r}
thunderstorm <- paste("THUDERSTORM","THUNDEERSTORM","THUNDERESTORM",
                           "THUNDERSTORM","TUNDERSTORM", "THUNDERTSORM",
                           "THUNDERSTROM","THUNDERTORM", "TSTM","THUNDESTORM",
                           "THUNERSTORM","GUSTNADO",
                           sep = "|")
high.temperature <- paste("HYPERTHERMIA","RECORD HIGH","RECORD TEMPERATURE",
                     "HOT","TEMPERATURE RECORD","HEAT",
                      sep = "|")
low.temperature <- paste("FREEZE","FROST","FREEZING","HYPOTHERMIA",
                        "RECORD LOW","LOW TEMPERATURE","LOW TEMP",
                        sep = "|")
high.swell <- paste("HIGH\\s*SWELLS","HEAVY SWELLS","TIDE",
                   "HIGH WAVE","HIGH WATER","COASTAL SURGE",
                   "HIGH SEA","WAVE","ROUGH SEA","HEAVY SEA",
                   sep = "|")
tornado  <- paste("TORNADO","TORNDAO","WATERSPOUT","WAYTERSPOUT","FUNNEL",
                  "WALL CLOUD","WATER SPOUT","LANDSPOUT",
                  sep = "|")
mapEventType <- function(x) {
    # order is important - eg: get Thunderstorm, Tropical storm first
    # before Storm
    if (grepl("BLIZZARD", x)) "BLIZZARD"
    else if (grepl(thunderstorm, x)) "THUNDERSTORM"
    else if (grepl(tornado, x)) "TORNADO/WATERSPOUT"
    else if (grepl("TROPICAL STORM|TYPHOON", x)) "TROPICAL STORM"
    else if (grepl("LIGHTING|LIGHTNING|LIGNTNING", x)) "LIGHTNING"
    else if (grepl("WIND|WND", x)) "WIND"
    else if (grepl("FLOOD|FLOOODING|STREAM|RISING WATER", x)) "FLOOD/SMALL STREAM"
    else if (grepl("AVALANCHE|AVALANCE", x)) "AVALANCHE"
    else if (grepl("COLD|COOL", x)) "COLD/COOL WEATHER"
    else if (grepl("HAIL", x)) "HAIL"   
    else if (grepl("SLIDE|SLUMP", x)) "LAND/MUD SLIDE"
    else if (grepl("RAIN|HEAVY SHOWER", x)) "RAIN/HEAVY SHOWER"  
    else if (grepl("SNOW", x)) "SNOW"   
    else if (grepl("HURRICANE|REMNANTS OF FLOYD", x)) "HURRICANE"
    else if (grepl("FIRE|RED FLAG CRITERIA", x)) "FIRE"
    else if (grepl("GLAZE|ICE|ICY|SLEET", x)) "GLAZE/ICE"
    else if (grepl(high.swell, x)) "HIGH SWELLS/TIDE/WAVE" 
    else if (grepl("WARM", x)) "WARM"     
    else if (grepl("WET", x)) "WET" 
    else if (grepl(low.temperature, x)) "FROST/FREEZE/HYPOTHERMIA"           
    else if (grepl(high.temperature, x)) "HIGH TEMPERATURE/HYPERTHERMIA"        
    else if (grepl("TSUNAMI", x)) "TSUNAMI"            
    else if (grepl("SMOKE", x)) "SMOKE"         
    else if (grepl("VOLCANIC", x)) "VOLCANIC"         
    else if (grepl("FOG|VOG", x)) "FOG" 
    else if (grepl("STORM", x)) "STORM"         
    else if (grepl("DRY|DRI|DROUGHT", x)) "DROUGHT/DRY"   
    else if (grepl("DAM", x)) "DAM BREAK/FAILURE"        
    else if (grepl("URBAN", x)) "URBAN/SMALL"        
    else if (grepl("WINTER|WINTRY", x)) "WINTER WEATHER"           
    else if (grepl("EROSION|EROSIN", x)) "EROSION "          
    else if (grepl("SURF|RIP CURRENT|DROWN", x)) "SURF/RIP CURRENT/DROWN"        
    else if (grepl("MARINE|SEICHE", x)) "MARINE ACCIDENT/SEICHE"  
    else if (grepl("PRECIPITATION|PRECIPATATION|PRECIP", x)) "PRECIPITATION"
    else if (grepl("MICROBURST|DOWNBURST", x)) "MICROBURST"  
    else if (grepl("DUST", x)) "DUST"         
    else if (grepl("TURBULENCE", x)) "TURBULENCE"           
    else "Other"
}
```

Damages are report in dollars and followed by an alphabetical character signifying the magnitude.  The [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) only specifies "K" for thousands, "M" for millions, and "B" for billions.  Other magnitudes are assumed; for example "H" for hundreds, and 
numeric values indicate the number of zeroes.
```{r}
toTenExponent <- function(x) {
    if (is.na(x)) return(0)
    
    numX <- as.numeric(x)
    if (!is.na(numX)) return(numX)

    upperX <- toupper(x)
    if (upperX == "H") return(2)
    else if (upperX == "K") return(3)
    else if (upperX == "M") return(6)
    else if (upperX == "B") return(9)
    else return(0)
}
```

Perform transformation
```{r}
data$newEVTYPE <- sapply(data$EVTYPE, mapEventType)
data$newEVTYPE <- as.factor(data$newEVTYPE)

data$HARMFUL <- data$FATALITIES + data$INJURIES

data$DMG <- data$PROPDMG * 10^sapply(data$PROPDMGEXP, toTenExponent) +
            data$CROPDMG * 10^sapply(data$CROPDMGEXP, toTenExponent)

event.harmful <- aggregate(HARMFUL ~ newEVTYPE, data = data, sum)
event.damage <- aggregate(DMG ~ newEVTYPE, data = data, sum)
```


## Results
Most harmful event - The bar chart below shows the total number fatalities and injuries
for each (mapped) category of weather events.
```{r}
g <- ggplot(event.harmful, aes(newEVTYPE, HARMFUL))
g <- g + geom_bar(colour="blue", stat="identity")
g <- g + ggtitle("Weather events harmful to the United States population")
g <- g + ylab("Number of injuries and fatalities")
g <- g + xlab("Event types")
g <- g + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1))
print(g)
```

As it can be seen, the most harmful event is:
```{r}
most.harmful <- event.harmful$HARMFUL == max(event.harmful$HARMFUL)
event.most.harmful <- event.harmful[most.harmful,]
event.most.harmful
```

Event causes the most damage - The bar chart below shows the total financial damage
of property and crop damanges for each (mapped) category of weather events.
```{r}
g <- ggplot(event.damage, aes(newEVTYPE, DMG / 1000000))
g <- g + geom_bar(colour="blue", stat="identity")
g <- g + ggtitle("Economic consequences of weather events across the United States")
g <- g + ylab("Damages in Millions of Dollars")
g <- g + xlab("Event types")
g <- g + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1))
print(g)
```

As it can be seen, the event that causes the most financial damage is:
```{r}
most.damage <- event.damage$DMG == max(event.damage$DMG)
event.most.damage <- event.damage[most.damage,]
event.most.damage
```